### 2022-08-23 article.tex
Use a semicolon instead of a plus before ggplot2::ggsave, as the plus now gives a rightful error

The example study followed is a pseudorandomly selected paper
from \cite{ahsan2017relative}. The primary data used by that paper is
from a population study called the Northern Swedish Population
Health Study (NSPHS) that started in 2010 \cite{igl2010northern}. 
The approximately 1000 participants were initially mostly surveyed
about lifestyle \cite{igl2010northern} and follow-up studies
provided the type of data relevant for this paper, 
which are (1) the genotypes \cite{johansson2013identification},
(2) the phenotypes, which are the concentrations of certain proteins
in the blood \cite{enroth2014strong,enroth2015effect}.

@@

The first type of primary data, the genotypes, 
consists out of single nucleotide polymorphisms (SNPs, pronounced as 'snips').
A SNP has a name and a location on the DNA, at which there is a certain nucleotide.
DNA (organized into chromosomes and present in every (nucleated) cell, 
see figure \ref{fig:eukakyote_dna}) 
consists out of billions of nucleotides.
There are four types of nucleotides, 
called adenosine, cytosine, guanine and thyrosine, all commonly abreviated
as A, C, G and T respectively.
One SNP example is \verb|rs12133641|, which is a SNP located at position 
154,428,283, where 67 percent of the people within this study have an A,
and 33 percent have a G (also from \cite{ahsan2017relative}, Table S3).

@@

The second type of data, the phenotypes, 
are concentrations of proteins in the blood. 
DNA contains the code for building proteins (see Figure 
\ref{fig:dna_to_protein}), 
as well as the rate
at which a protein is created. Some proteins end up in the blood and
their presence can be used to assess the health of an individual.
IL6RA is one such protein and, spoiler alert, its concentration
is associated with the SNP mentioned earlier.

The field of genetic epidemiology looks -among others- for
correlations between genetic data and biological traits.
For example, figure \ref{fig:ahsan2017relative_table_2_sub} 
(from \cite{ahsan2017relative}) shows that
SNP \verb|rs12133641| is highly correlated (p-value is $3.0^{-73}$, 
for $n$ = 961 individuals) with protein IL6RA. What this results does
not yet teach us, is how this correlation works, yet
figure \ref{fig:ahsan2017relative_s6} shows us the direction of the association:
the X axis shows the possible genetic make-ups (aka the genotype) of the individuals,
where 0 denotes the individuals with genotype
AA (the individual inherited one adenosine 
from his/here mother and one adenosine from his/her father), 
1.0 denotes AG (one A is inherited from one parent, 
where the G is inherited from the other parent) and 2.0 denotes GG.
From Figure \ref{fig:ahsan2017relative_s6} we can conclude that, on average,
the more guanines are present at that SNPs location,
the higher concentration of IL6RA can be found in a human's blood.

The amount of variance that can be explained by an association (i.e.
the R squared value) is rarely 100 percent, which means that a trait (in
this case, the concentration of IL6RA) cannot be perfectly explained
from the genotype alone. As we can see in figure \ref{fig:ahsan2017relative_s6}, 
43 percent of the variance can be attributed to an individuals'
genotype. 
Additional factors, 
such as the effect
of the environment (e.g. geographic location, time of day), 
lifestyle (e.g. smoking yes/no) or having a disease (e.g. diabetes) 
are needed to explain the additional variation.

@@

These conclusions may end up in the clinic.
For the sake of having a clear example,
let's assume that a high level of IL6RA 
is associated with a disease that develops later in life,
yet is preventable by lifestyle changes.
Would this be the case, we can create a gene panel that specifically measures
SNP \verb|rs12133641|. 
If the gene panel shows an individual has two guanines, 
we know that this person is likelier to develop higher levels of
IL6RA and is likelier to benefit from the lifestyle changes.

In this simple example, it will be easier
to measure the level of IL6RA in the blood, than using a gene panel.
However, there are associations published for many diseases,
in which multiple SNPs contribute to being more likely 
to develope a disease in the future. Here, the phenotype (having 
a disease in the future) is impossible to detect at the present
and associations found in earlier studies are used to create a gene panel.
As creating a gene panel is costly, those associations better be correct.

>>>>

The example study followed is a pseudorandomly selected paper
from \cite{ahsan2017relative}. The primary data used by that paper is
from a population study called the Northern Swedish Population
Health Study (NSPHS) that started in 2010 \cite{igl2010northern}. 
The approximately 1000 participants were initially mostly surveyed
about lifestyle \cite{igl2010northern} and follow-up studies
provided the type of data relevant for this paper, 
which are (1) the genotypes \cite{johansson2013identification},
(2) the phenotypes, in this case, concentrations of certain proteins in the blood \cite{enroth2014strong,enroth2015effect}.

@@

The first type of primary data, the genotypes, 
consists out of single nucleotide polymorphisms (SNPs, pronounced as 'snips').
A SNP has a name and a location on the DNA, at which there is a certain nucleotide.
DNA (organized into chromosomes and present in every (nucleated) cell, 
\iffalse
(see figure \ref{fig:eukakyote_dna}))
\fi
consists out of billions of nucleotides.
There are four types of nucleotides, 
called adenosine, cytosine, guanine and thyrosine, all commonly abreviated
as A, C, G and T respectively.
One SNP example is \verb|rs12133641|, which is a SNP located at position 
154,428,283, where 67 percent of the people within this study have an A,
and 33 percent have a G (also from \cite{ahsan2017relative}, Table S3).

@@

The second type of data, the phenotypes, 
are concentrations of proteins in the blood. 
DNA contains the code for building proteins
\iffalse
(see Figure \ref{fig:dna_to_protein})
\fi
, as well as the rate
at which a protein is created. Some proteins end up in the blood and
their presence can be used to assess the health of an individual.
IL6RA is one such protein and, spoiler alert, its concentration
is associated with the SNP mentioned earlier.

The field of genetic epidemiology looks -among others- for
correlations between genetic data and biological traits.
For example, \cite{ahsan2017relative} shows that
SNP \verb|rs12133641| is highly correlated (p-value is $3.0^{-73}$
\iffalse
(see figure \ref{fig:ahsan2017relative_table_2_sub})
\fi
,
for $n$ = 961 individuals) with protein IL6RA.
The direction of the association is also concluded:
the more guanines (as opposited to adenines) are present at that SNPs location,
the higher concentration of IL6RA can be found in a human's blood
\iffalse
(see figure \ref{fig:ahsan2017relative_table_2_sub})
\fi
. The amount of variance that can be explained by an association (i.e.
the R squared value) is rarely 100 percent, which means that a trait (in
this case, the concentration of IL6RA) cannot be perfectly explained
from the genotype (in this case, SNP \verb|rs12133641|) alone. 
In this example, 43 percent of the variance 
can be attributed to an individuals' genotype
\iffalse
(see figure \ref{fig:ahsan2017relative_table_2_sub})
\fi
. Additional factors, 
such as the effect
of the environment (e.g. geographic location, time of day), 
lifestyle (e.g. smoking yes/no) or having a disease (e.g. diabetes) 
are needed to explain the additional variation.

@@

The conclusions drawn from this paper,
may end up in the clinic.
For the sake of having a clear (yet fictuous) example,
let's assume that a high level of IL6RA 
is associated with a disease that develops later in life,
yet is preventable by lifestyle changes.
Would this be the case, we can create a gene panel that specifically measures
SNP \verb|rs12133641|. 
If the gene panel shows an individual has two guanines, 
we know that this person is likelier to develop higher levels of
IL6RA and is likelier to benefit from the lifestyle changes.

From this simple example, it will be easier
to measure the level of IL6RA in the blood, than using a gene panel.
However, there are associations published for many diseases,
in which multiple SNPs contribute to being more likely 
to develope a disease in the future. Here, the phenotype (having 
a disease in the future) is impossible to detect at the present
and associations found in earlier studies are used to create a gene panel.
As creating a gene panel is costly, those associations better be correct.
